\pkg{Rgbp}: An \proglang{R} Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data

### 2015-07-31 jss/article_v6.tex
v6

\section[Hierarchical Structure]{Conjugate hierarchical modeling structure} \label{sec2}
%A two-level  multilevel model is a powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital-level data in a certain state.

@@

\subsection{Estimation for random effects}
It is intractable to derive analytically the unconditional posterior distribution of each random effect  for the three models. This motivates us to use the ADM again by assuming that each random effect has a skewed Normal distribution \citep{azzalini1985class} for the Gaussian model, a Beta distribution for the Binomial model, or a Gamma distribution for the Poisson model; for $j=1, 2, \ldots, k$,
\begin{align}
\mu_j\vert\boldsymbol{y} &\sim \textrm{skewed-Normal}(\phi,~ \omega,~ \delta),\label{skewnormal}\\
p_j\vert \boldsymbol{y} &\sim \textrm{Beta}(t_{1j},~ t_{0j}),\label{betapost}\\
\lambda_j\vert \boldsymbol{y} &\sim \textrm{Gamma}(s_{1j},~ s_{0j}),\label{gammapost}
\end{align}
where $(\phi, \omega, \delta)$ of the skewed-Normal distribution  are location,  scale, and skewness parameters, respectively.

\cite{morris2012} first note that the unconditional posterior distribution of the random effect in a two-level conjugate Gaussian model may be skewed. \cite{kelly2014advances}  shows that the skewed-Normal approximation to the unconditional posterior distribution of the random effect is better than a Normal approximation ($\mu_j\vert\boldsymbol{y}\sim \textrm{Normal}$) in terms of the repeated sampling coverage properties of random effects. \cite{kelly2014advances} uses  the results derived in \cite{azzalini1985class}, i.e., $\E(\mu_j\vert \boldsymbol{y}) =\phi+\omega\delta\sqrt{2/\pi}$, Var$(\mu_j\vert \boldsymbol{y})=\omega^2(1-2\delta^2/\pi)$, and Skewness$(\mu_j\vert \boldsymbol{y})=(4-\pi)\delta^3/[2(\pi/2-\delta^2)^{3/2}]$. He estimates three parameters of the skewed-Normal distribution using  three approximately estimated unconditional posterior  moments of each random effect; see \cite{kelly2014advances} for details.

@@

The S3 object ``gbp'' benefits from three generic functions, \code{print}, \code{summary}, and \code{plot}. The estimation result for all the random effects appears if we type the ``gbp'' object at the \proglang{R} console, which plays the same role of the function \code{print} with its default argument ``\code{sort = TRUE}''. When the argument \code{sort} is set to \code{TRUE}, the function \code{print} prints out the estimation result for all the groups in the ascending order of $n$ for the Binomial and Poisson model and the descending order of standard errors for the Gaussian model. When the argument \code{sort} is \code{FALSE}, the estimation result comes out in the order of data input.
\begin{CodeChunk}
\begin{CodeInput}
R> b.output
R> print(b.output, sort = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Frequency method checking}
The function \code{coverage} conducts a frequency method checking that estimates the coverage probabilities of  random effects, conditioning on the values of the hyper-parameters that generate random effects and  mock data sets. The basic usage of \code{coverage} needs a ``gbp'' object, such as \code{b.output} above, as the first argument;
\begin{CodeChunk}
\begin{CodeInput}
R> cov <- coverage(b.output, nsim = 1000)
\end{CodeInput}
\end{CodeChunk}

@@

These data can be loaded into \proglang{R} using the following code.
\begin{CodeChunk}
\begin{CodeInput}
R> library("Rgbp")
R> data("hospital")
R> y <- hospital$d
R> n <- hospital$n
\end{CodeInput}
\end{CodeChunk}

@@

The function \code{gbp} fits the Poisson model with a Gamma conjugate prior distribution as the population distribution of the death rates in New York states whose mean ($\lambda^E$) is 0.030. The number of regression coefficients ($m$) is 0 because we do not estimate them via  a log-linear regression. 
\begin{CodeChunk}
\begin{CodeInput}
R> p.output <- gbp(z, n, mean.PriorDist = 0.03, model = "poisson")
R> p.output
\end{CodeInput}
\begin{CodeOutput}
Summary for each unit (sorted by n):

@@

The Education Testing Service (ETS) conducted randomized experiments in eight separate schools (groups) to test whether students (units) SAT scores are effected by coaching. The dataset contains the estimated coaching effects on SAT scores ($y_{j}, j=1, \ldots, 8$) and standard errors ($se_{j}, j=1, \ldots, 8$) of the eight schools \citep{1981}. We can load this data set into \proglang{R} by the following codes.
\begin{CodeChunk}
\begin{CodeInput}
R> library("Rgbp")
R> data("schools")
R> y  <- schools$y
R> se <- schools$se
\end{CodeInput}
\end{CodeChunk}

@@

The data of 18 major league baseball players contain the batting averages through their first 45 official at-bats of the 1970 season \citep{1975}. We add a binary covariate that takes on one if a player is an outfielder and zero otherwise. We can load the data into \proglang{R} with the following codes.
\begin{CodeChunk}
\begin{CodeInput}
R> library("Rgbp")
R> data("baseball")
R> y <- baseball$Hits
R> n <- baseball$At.Bats
R> x <- ifelse(baseball$Position == "fielder", 1, 0)
\end{CodeInput}
\end{CodeChunk}
%$
Conditioning on the unknown true batting average (random effect) of each player we assume that the at-bats are independent and therefore, $y_{j}\vert p_{j}\stackrel{\textrm{indep.}}{\sim} \textrm{Binomial}(45,~ p_{j}), ~j=1, \ldots, 18$. Our goal is to obtain point and interval estimates of each random effect whilst considering the additional information on whether the player is an outfielder or not. The function \code{gbp} provides a way to incorporate such covariate information seamlessly into the model so that the regression towards the mean (RTTM) occurs within outfielders and non-outfielders separately. %For reference, we can also assume the Normal distribution using $y_{j}=z_{j} / n_{j}$ and $V_{j}=\bar{y}(1-\bar{y})/n$, where $\bar{y}=\sum_{j} z_{j} / \sum_{j} n_{j}$, $j=1,\ldots, 18$.

@@

If there are at least two groups whose observed values $y_j$ are non-zero, then $f(r\vert\boldsymbol{y})$ goes to zero as $r$ goes to zero due to the polynomial function of $r$ in \eqref{final_post_proper}; the following two factors in \eqref{final_post_proper} approach one. As $r$ becomes infinite, $f(r\vert\boldsymbol{y})$ touches zero exponentially fast due to the exponential term in the middle of \eqref{final_post_proper}. Thus, the integration of $f(r\vert\boldsymbol{y})$ must be finite.

@@

>>>>

@@

\subsection{Estimation for random effects}
It is intractable to derive analytically the unconditional posterior
distribution of each random effect for the three models. Thus we approximate the
distributions by matching the estimated posterior moments with a skewed-Normal distribution \citep{azzalini1985class} for
the Gaussian model, a Beta distribution for the Binomial model and a Gamma distribution for the Poisson model; for $j=1, 2, \ldots, k$,
\begin{align}
\mu_j\vert\boldsymbol{y} &\sim \textrm{skewed-Normal}(\phi,~ \omega,~ \delta),\label{skewnormal}\\
p_j\vert \boldsymbol{y} &\sim \textrm{Beta}(t_{1j},~ t_{0j}),\label{betapost}\\
\lambda_j\vert \boldsymbol{y} &\sim \textrm{Gamma}(s_{1j},~ s_{0j}),\label{gammapost}
\end{align}
where $(\phi, \omega, \delta)$ of the skewed-Normal distribution  are location,  scale, and skewness parameters, respectively.

\cite{morris2012} first note that the unconditional posterior distribution of
the random effect in a two-level conjugate Gaussian model may be
skewed. \cite{kelly2014advances}  shows that the skewed-Normal approximation to
the unconditional posterior distribution of the random effect is better than a
Normal approximation ($\mu_j\vert\boldsymbol{y}\sim \textrm{Normal}$) in terms
of the repeated sampling coverage properties of random
effects. \cite{kelly2014advances} estimates the first three moments of
the random-effects by noting that $\mu_j | \mathbf{y}, B_{j}$ is normally
distributed and thus estimates the moments by using the ADM approximation of
the shrinkage factors, $B_{j}$, and the law of third cumulants \cite{brill}. The
three estimated moments are then matched to the first three moments of the
skewed-Normal distribution, $\E(\mu_j\vert \boldsymbol{y})
=\phi+\omega\delta\sqrt{2/\pi}$, Var$(\mu_j\vert
\boldsymbol{y})=\omega^2(1-2\delta^2/\pi)$, and Skewness$(\mu_j\vert
\boldsymbol{y})=(4-\pi)\delta^3/[2(\pi/2-\delta^2)^{3/2}]$
\cite{azzalini1985class}. The full derivation can be found in \cite{kelly2014advances}.

@@

The S3 object ``gbp'' benefits from three generic functions, \code{print},
\code{summary}, and \code{plot}. The estimation result for all the random
effects appears if we type the ``gbp'' object at the \proglang{R} console, which
plays the same role of the function \code{print} with its default argument
``\code{sort = TRUE}''. When the argument \code{sort} is set to \code{TRUE}, the
function \code{print} prints out the estimation result for all the groups in the
ascending order of $n$ for the Binomial and Poisson model and the descending
order of standard errors for the Gaussian model. When the argument \code{sort}
is \code{FALSE}, the estimation result is returned in the order of data input.
\begin{CodeChunk}
\begin{CodeInput}
R> b.output
R> print(b.output, sort = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Frequency method checking}
The function \code{coverage} conducts frequency method checking. It estimates
the coverage properties for our estimators of the random effects at
a particular value of the hyperparameters by averaging the coverage over many simulated datasets. The basic usage of \code{coverage} needs a ``gbp'' object, such as \code{b.output} above, as the first argument;
\begin{CodeChunk}
\begin{CodeInput}
R> cov <- coverage(b.output, nsim = 1000)
\end{CodeInput}
\end{CodeChunk}

@@

The following code
\begin{CodeChunk}
\begin{CodeInput}
R> library("Rgbp")
R> data("hospital")
R> y <- hospital$d
R> n <- hospital$n
\end{CodeInput}
\end{CodeChunk}
can be used to load these data into \proglang{R}.

@@

The function \code{gbp} can then be used to fit a Poisson-Gamma to the death
rates in New York states with prior mean, ($\lambda^E$), equal to 0.030 and the
number of regression coefficients ($m$) set to 0.

\begin{CodeChunk}
\begin{CodeInput}
R> p.output <- gbp(z, n, mean.PriorDist = 0.03, model = "poisson")
R> p.output
\end{CodeInput}
\begin{CodeOutput}
Summary for each unit (sorted by n):

@@

The Education Testing Service (ETS) conducted randomized experiments in eight
separate schools (groups) to test whether students (units) SAT scores are
effected by coaching. The dataset contains the estimated coaching effects on SAT
scores ($y_{j}, j=1, \ldots, 8$) and standard errors ($se_{j}, j=1, \ldots, 8$)
of the eight schools \citep{1981}. These data are contained in the package and can be loaded into \proglang{R}
\begin{CodeChunk}
\begin{CodeInput}
R> library("Rgbp")
R> data("schools")
R> y  <- schools$y
R> se <- schools$se
\end{CodeInput}
\end{CodeChunk}

@@

The data of 18 major league baseball players contain the batting averages
through their first 45 official at-bats of the 1970 season \citep{1975}. A
binary covariate is created that takes on the value one if a player is an
outfielder and zero otherwise. The data can be loaded into \proglang{R} with the following code
\begin{CodeChunk}
\begin{CodeInput}
R> library("Rgbp")
R> data("baseball")
R> y <- baseball$Hits
R> n <- baseball$At.Bats
R> x <- ifelse(baseball$Position == "fielder", 1, 0)
\end{CodeInput}
\end{CodeChunk}
%$
Conditional on the unknown true batting average (random effect) of each player
it is assumed that the at-bats are independent and therefore, $y_{j}\vert p_{j}\stackrel{\textrm{indep.}}{\sim} \textrm{Binomial}(45,~ p_{j}), ~j=1, \ldots, 18$. Our goal is to obtain point and interval estimates of each random effect whilst considering the additional information on whether the player is an outfielder or not. The function \code{gbp} provides a way to incorporate such covariate information seamlessly into the model so that the regression towards the mean (RTTM) occurs within outfielders and non-outfielders separately. %For reference, we can also assume the Normal distribution using $y_{j}=z_{j} / n_{j}$ and $V_{j}=\bar{y}(1-\bar{y})/n$, where $\bar{y}=\sum_{j} z_{j} / \sum_{j} n_{j}$, $j=1,\ldots, 18$.

@@

If there are at least two groups whose observed values $y_j$ are non-zero, then $f(r\vert\boldsymbol{y})$ goes to zero as $r$ goes to 0 due to the polynomial function of $r$ in \eqref{final_post_proper}; the following two factors in \eqref{final_post_proper} approach one. As $r$ becomes infinite, $f(r\vert\boldsymbol{y})$ touches zero exponentially fast due to the exponential term in the middle of \eqref{final_post_proper}. Thus, the integration of $f(r\vert\boldsymbol{y})$ must be finite.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
